---
title: "Video Quiz: Foundations of Logistic Regression"
subtitle: "Prelude to Module 2, Steps 1-1 through 2-1"
author: "Ron Sabilla"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: right-body
    toc-title: "Questions"
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: paged
    link-external-newwindow: true
    embed-resources: true
execute:
  warning: false
  message: false
  freeze: true
---

# Overview {.unnumbered}

This quiz covers four StatQuest videos that build the conceptual
foundation for logistic regression — the model I fitted in the Week 2
notebook. Below I've reframed each prompt as the question I'm
personally working through after watching each video, followed by my
code and response.

| Step | Video | Link |
|--------------|---------------------------|-------------------------------------|
| 1-1 | Logs (Logarithms), Clearly Explained!!! | <https://www.youtube.com/watch?v=VSi0Z04fWj0> |
| 1-2 | Odds and Log(Odds), Clearly Explained!!! | <https://www.youtube.com/watch?v=ARfXDSkQf1Y> |
| 1-3 | Odds Ratios and Log(Odds Ratios), Clearly Explained!!! | <https://www.youtube.com/watch?v=8nm0G-1uJzA> |
| 2-1 | Logistic Regression, Clearly Explained!!! | <https://www.youtube.com/watch?v=yIYKR4sgzI8> |

------------------------------------------------------------------------

```{r setup}
library(tidyverse)

# Recreate the direct-mail dataset used throughout the course
set.seed(123)
n <- 3000

mail_data <- tibble(
  customer_id = paste0("C", str_pad(1:n, 5, pad = "0")),
  age = round(rnorm(n, mean = 45, sd = 12)),
  income = round(rlnorm(n, meanlog = 10.8, sdlog = 0.6)),
  recency_days = round(rexp(n, rate = 1 / 60)),
  freq_12mo = rpois(n, lambda = 3),
  avg_order_amt = round(rlnorm(n, meanlog = 4.2, sdlog = 0.5), 2),
  channel = sample(
    c("email", "direct_mail", "digital"),
    n,
    replace = TRUE,
    prob = c(0.5, 0.3, 0.2)
  ),
  region = sample(
    c("West", "South", "Midwest", "Northeast"),
    n,
    replace = TRUE
  ),
  loyalty_tier = sample(
    c("Bronze", "Silver", "Gold", "Platinum"),
    n,
    replace = TRUE,
    prob = c(0.4, 0.3, 0.2, 0.1)
  ),
  responded = rbinom(
    n,
    1,
    prob = plogis(
      -3 +
        0.02 * (age - 45) +
        0.3 * log(income / 50000) +
        0.1 * freq_12mo -
        0.005 * recency_days
    )
  )
) |>
  mutate(
    income = if_else(runif(n) < 0.08, NA_real_, income),
    avg_order_amt = if_else(runif(n) < 0.05, NA_real_, avg_order_amt),
    responded = factor(responded, levels = c(1, 0), labels = c("yes", "no"))
  )
```

------------------------------------------------------------------------

# Step 1-1 — Logarithms {#sec-logs}

> **Video:** Logs (Logarithms), Clearly Explained!!! <https://www.youtube.com/watch?v=VSi0Z04fWj0>

## Q1 — What Does a Log Do?

**My question:** the video says a log's core job is to isolate the
exponent — does that actually check out in R, and why would that
matter for a dataset where income swings from \$7,000 to \$455,000?

```{r q1-1}
log(32, base = 2)
```

**Response:** $\log_2(32) = 5$ because $2^5 = 32$ — the log answers
"what exponent do I raise the base to, to get this number?" R
confirms it: `log(32, base = 2)` returns `5`. This is useful for a
variable like income that spans two orders of magnitude, because the
log compresses that huge range down to a much smaller, more even
scale — the jump from \$7K to \$70K (a 10x change) looks the same size
on the log scale as the jump from \$70K to \$700K, instead of the raw
dollar gaps looking wildly different in size.

## Q2 — Multiplication Becomes Addition

```{r q1-2}
log(4 * 8, base = 2)
log(4, base = 2) + log(8, base = 2)
```

**Response:** Both sides equal `5`, confirming
$\log(a \times b) = \log(a) + \log(b)$. In logistic regression, we're
implicitly multiplying many small probabilities together when
maximizing likelihood — repeated multiplication of numbers less than
1 shrinks toward zero fast and can underflow computer precision.
Converting that product into a **sum of logs** keeps the numbers in a
stable, workable range, which is exactly why models optimize
*log*-likelihood rather than raw likelihood.

## Q3 — Geometric Mean in Marketing

```{r q1-3}
orders <- c(10, 20, 80, 40, 1000)
mean(orders)
exp(mean(log(orders)))
```

**Response:** The arithmetic mean is \$230, while the geometric mean
is about \$57.71. The geometric mean better represents the "typical"
customer here because four of the five orders are between \$10 and
\$80 — the \$1,000 order is a single outlier that drags the arithmetic
mean up more than triple the next-highest order. The geometric mean
works in multiplicative space, so it isn't thrown off by one extreme
value the way an additive average is.

## Q4 — Log(0) and the Offset

```{r q1-4}
log(0)
log(0 + 1)
```

**Response:** `log(0)` returns `-Inf` since there's no exponent you
can raise a base to that produces zero. My Week 2 recipe uses
`step_log(..., offset = 1)` because several columns (like
`avg_order_amt` or `recency_days`) could plausibly contain a `0`, and
logging a true zero would break the pipeline with `-Inf`. Adding 1
before logging (`log1p`-style) is a safe fix because `log(0 + 1) = 0`
— it keeps zero mapped to a real, finite number instead of infinity,
while barely affecting larger values.

## Q5 — Base Conversion

```{r q1-5}
income_val <- 49021
log(income_val, base = 2)
log10(income_val)
log(income_val)

# Ratio
log(income_val) / log10(income_val)
```

**Response:** For income = \$49,021: `log2()` gives about 15.58,
`log10()` gives about 4.69, and the natural log gives about 10.8.
Dividing the natural log by `log10()` gives a constant
`r round(log(10), 4)`, which is `log(10)` — confirming all three bases
differ only by a fixed multiplier. After `step_normalize()` centers
and scales the variable, that constant multiplier gets fully absorbed
into the scaling step, so it does **not** matter which base I used
for `step_log()` — the final standardized values come out identical
either way.

------------------------------------------------------------------------

# Step 1-2 — Odds and Log(Odds) {#sec-odds}

> **Video:** Odds and Log(Odds), Clearly Explained!!! <https://www.youtube.com/watch?v=ARfXDSkQf1Y>

## Q6 — Probability vs. Odds

```{r q2-1}
yes <- sum(mail_data$responded == "yes")
no <- sum(mail_data$responded == "no")

# Probability
p <- yes / (yes + no)
p

# Odds
odds <- yes / no
odds
```

**Response:** 142 out of 3,000 customers responded, giving a
probability of about 4.73% and odds of about 0.0497. In plain
English, odds of 0.05 mean that for roughly every 1 customer who
responds, about 20 do not — a marketer could read this as "for every
20 non-responders, expect 1 responder." It's a ratio of two group
sizes rather than a share of the whole, which is why it feels less
intuitive than a straight percentage but is what logistic regression
actually models under the hood.

## Q7 — Converting Between Probability and Odds

```{r q2-2}
prob_to_odds <- function(p) p / (1 - p)

prob_to_odds(0.20)
prob_to_odds(0.50)
prob_to_odds(0.80)
```

**Response:** A 20% chance gives odds of 0.25 (1-to-4), a 50% chance
gives odds of exactly 1 (even odds), and an 80% chance gives odds of
4 (4-to-1 in favor). As probability climbs toward 1, odds shoot up
toward infinity — the relationship isn't linear. This asymmetry (odds
compress between 0 and 1 below the 50% mark, then stretch out to
infinity above it) is exactly the problem the log transform in Q8
fixes.

## Q8 — Why Log(Odds)?

```{r q2-3}
probs <- c(0.10, 0.50, 0.90)
odds8 <- probs / (1 - probs)
log_odds8 <- log(odds8)

tibble(probability = probs, odds = odds8, log_odds = log_odds8)
```

**Response:** At $P = 0.50$, `log(odds)` comes out to exactly 0,
which makes intuitive sense — 50/50 is the point of "no lean" in
either direction, so it should sit at the center of a symmetric
scale. The log(odds) values for 0.10 and 0.90 come out as
$\mp 2.197$ — perfectly symmetric around zero. That symmetry is the
whole point for modeling: it turns odds (which were lopsided, ranging
0 to 1 on one side and 1 to infinity on the other) into a scale that
behaves like a normal, unbounded number line — which is exactly what
a linear model needs to work with.

## Q9 — The Logit Function

```{r q2-4}
# Overall response rate
p_respond <- mean(mail_data$responded == "yes")
p_respond

# Convert to log(odds)
log(p_respond / (1 - p_respond))
```

**Response:** The overall response rate is about 4.73%, which
converts to a log(odds) of about **−3.00** — this is essentially the
same number as the `-3` intercept baked into the `plogis()` call in
the setup chunk, which makes sense since that intercept *is* the
baseline log-odds of responding before any predictor effects are
added. To convert a predicted probability like `.pred_yes` back to
log(odds), I'd write: `log(preds$.pred_yes / (1 - preds$.pred_yes))`.

## Q10 — Log(Odds) and the Normal Distribution

**Response:** Income is right-skewed because a small number of very
high earners stretch the tail, so logging it pulls those extremes
back toward a roughly bell-shaped distribution that's easier to model
linearly. Log(odds) of a binary outcome behaves similarly — because
it's unbounded and roughly symmetric rather than squeezed into [0,1]
like probability, it's a much more natural target for a **linear**
combination of predictors. That's the whole premise of logistic
regression: instead of trying to draw a straight line through raw
0/1 or probability values (which breaks down at the edges), we draw
the straight line through log(odds) space, where a linear model
actually behaves well, and then convert back to probability at the
end with `plogis()`.

------------------------------------------------------------------------

# Step 1-3 — Odds Ratios and Log(Odds Ratios) {#sec-or}

> **Video:** Odds Ratios and Log(Odds Ratios), Clearly Explained!!! <https://www.youtube.com/watch?v=8nm0G-1uJzA>

## Q11 — What Is an Odds Ratio?

```{r q3-1}
tier_summary <- mail_data |>
  group_by(loyalty_tier) |>
  summarise(
    yes = sum(responded == "yes"),
    no = sum(responded == "no"),
    odds = yes / no
  )

tier_summary

# Odds ratio: Gold vs Bronze
gold_odds <- tier_summary |> filter(loyalty_tier == "Gold") |> pull(odds)
bronze_odds <- tier_summary |> filter(loyalty_tier == "Bronze") |> pull(odds)

OR <- gold_odds / bronze_odds
OR
```

**Response:** Gold customers have odds of responding of about 0.0500,
and Bronze customers about 0.0507 — almost identical. The odds ratio
(Gold ÷ Bronze) comes out to about 0.985, just barely below 1. Since
the whole point of an OR is distance from 1 in either direction,
Gold tier is actually a **very weak** predictor of response relative
to Bronze in this sample — practically no difference at all, even
though I might have expected loyalty tier to matter more.

## Q12 — Asymmetry and Log(Odds Ratio)

```{r q3-2}
# Log odds ratio
log(OR)

# Reverse direction
OR_reverse <- bronze_odds / gold_odds
log(OR_reverse)
```

**Response:** `log(OR)` is about −0.0149 (very slightly negative,
consistent with the OR being just under 1). Flipping the comparison
direction (Bronze ÷ Gold) gives the reciprocal OR, and its log comes
out to about **+0.0149** — the exact same magnitude, just flipped in
sign. That's the payoff of working in log space: reversing the
comparison direction only flips the sign instead of transforming the
number in some more complicated way, which wouldn't be true for the
raw (non-logged) odds ratio. Log(OR) = 0 is the natural reference
point for "no relationship" because it's the log of an OR of exactly
1 — meaning the two groups' odds are identical.

## Q13 — Effect Size Interpretation

```{r q3-3}
tier_summary <- tier_summary |>
  mutate(OR_vs_bronze = odds / bronze_odds, log_OR = log(OR_vs_bronze)) |>
  arrange(desc(abs(log_OR)))

tier_summary
```

**Response:** Platinum has the strongest association with response
(OR ≈ 1.48 vs. Bronze, log(OR) ≈ 0.394 — the largest magnitude of any
tier), while Gold has the weakest (log(OR) ≈ −0.015, essentially no
difference from Bronze). In one sentence for a non-technical
manager: "Platinum members are about 48% more likely, in odds terms,
to respond to this campaign than our Bronze baseline, making them our
strongest audience for this type of promotion."

## Q14 — Fisher's Exact Test

```{r q3-4}
gold_bronze <- mail_data |>
  filter(loyalty_tier %in% c("Gold", "Bronze"))

ct <- table(gold_bronze$loyalty_tier, gold_bronze$responded)
ct

fisher.test(ct)
```

**Response:** Fisher's Exact Test on Gold vs. Bronze returns a
p-value of essentially 1 — nowhere close to the 0.05 threshold. That
tells me the small difference in response odds between Gold and
Bronze customers is entirely explainable by random chance; there's no
statistically significant relationship between being a Gold vs.
Bronze member and responding. I'd prefer Fisher's Exact Test over
chi-square whenever any cell in the contingency table is small (as a
rule of thumb, expected counts under 5), since Fisher's calculates
the exact probability rather than relying on a large-sample
approximation.

## Q15 — Chi-Square Test

```{r q3-5}
chisq.test(ct)
```

**Response:** The chi-square test on the same table also returns a
p-value of essentially 1, matching Fisher's conclusion exactly — no
significant relationship. With 3,000 rows (and reasonably large cell
counts within the Gold/Bronze subset, both well above 5), chi-square
is the more appropriate everyday choice here since it's
computationally lighter and built for larger samples; Fisher's exact
test is really meant as a safeguard for smaller or sparser tables.

## Q16 — Wald Test Concept

**Response:**

1. A log(odds ratio) being "many standard deviations from zero" means
   the estimated effect is large relative to how much noise or
   uncertainty surrounds that estimate — in other words, it's
   unlikely we'd see an effect this far from "no relationship" (0) if
   the true effect were actually zero.
2. Lasso's penalty and the Wald test are aiming at the same goal from
   different directions: Wald asks "is this coefficient statistically
   distinguishable from zero?" after the fact, while Lasso directly
   shrinks weak, noisy coefficients *to* exactly zero during fitting.
   Both are ways of separating predictors that carry real signal from
   ones that don't.
3. The video's practical guideline is that there's no universally
   "correct" test — Fisher's, chi-square, and Wald all approximate
   the same underlying question, so the practical advice is to pick
   the test suited to your sample size and data structure rather than
   searching for one "right" answer.

## Q17 — Connecting Odds Ratios to the Course

**Response:**

1. If `freq_12mo`'s coefficient is positive, it implies that as a
   customer's purchase frequency increases, the **odds** of them
   responding to the campaign increase as well — each one-unit
   increase in `freq_12mo` multiplies the odds of responding by
   $e^{\beta}$.
2. After `step_normalize()`, every numeric predictor is on the same
   standardized (mean 0, sd 1) scale, so a coefficient of, say, 0.3
   means the same thing regardless of whether it came from income
   (originally in tens of thousands of dollars) or frequency
   (originally a small count). That makes it fair to directly compare
   coefficient magnitudes across predictors to see which one has the
   strongest relationship with the odds of responding.

------------------------------------------------------------------------

# Step 2-1 — Logistic Regression {#sec-logit}

> **Video:** Logistic Regression, Clearly Explained!!! <https://www.youtube.com/watch?v=yIYKR4sgzI8>

## Q18 — Linear vs. Logistic Regression

**Response:**

1. Linear regression predicts a continuous numeric value; logistic
   regression predicts a **probability** (bounded between 0 and 1)
   for a categorical, typically binary, outcome.
2. A straight line has no ceiling or floor, so for a yes/no outcome
   it will happily predict values below 0 or above 1 for extreme
   predictor values — which is nonsensical for something that's
   supposed to represent "will respond" vs. "won't respond."
3. With a ~4.7% baseline response rate, a linear regression fit to a
   very frequent, high-income customer could easily predict a value
   well above 1 (say, 1.3), which is meaningless as a probability —
   there's no such thing as a "130% chance of responding."

## Q19 — The S-Shaped Logistic Curve

$$P = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x)}}$$

```{r q4-2}
plogis(-3) # very negative linear predictor
plogis(0) # at zero
plogis(3) # very positive linear predictor
```

**Response:** `plogis(0)` returns exactly `0.5`, which makes sense
from the formula — when the linear predictor inside the exponent is
0, $e^{0} = 1$, so $P = 1/(1+1) = 0.5$: the exact midpoint of the
curve. Looking at the setup chunk, the `responded` column was
generated from
`plogis(-3 + 0.02*(age-45) + 0.3*log(income/50000) + 0.1*freq_12mo - 0.005*recency_days)`.
In marketing terms, that linear predictor represents a customer's
combined "propensity score" before converting to probability — the
−3 is the baseline log-odds for an average customer, and each term
after it nudges that score up or down based on how much older, wealthier,
more frequent a purchaser, or more recently active than average that
customer is.

## Q20 — Maximum Likelihood vs. Least Squares

**Response:**

1. Logistic regression can't use least squares because the
   "residual" for a 0/1 outcome compared to a predicted *probability*
   doesn't behave like a normal, symmetric error term — the variance
   of a binary outcome depends on the predicted probability itself,
   so squared-error residuals don't have the constant-variance
   property least squares relies on.
2. "Maximizing the likelihood" means logistic regression searches for
   the S-curve (the combination of coefficients) that makes the
   observed pattern of yeses and nos in the data as probable as
   possible — rather than minimizing distances to a line, it's
   asking "which curve makes what I actually saw most likely to have
   happened?"
3. Since there's no traditional residual, model fit for logistic
   regression is assessed with classification-specific metrics
   instead — things like accuracy, sensitivity/specificity, ROC-AUC,
   and log-loss (metrics I'll compute with `yardstick` in the Week 3
   notebook), rather than R² or residual plots.

## Q21 — Predicting Probability vs. Class

```{r q4-4}
#| eval: false
library(tidymodels)
tidymodels_prefer()

mail_split <- initial_split(mail_data, prop = 0.80, strata = responded)
mail_train <- training(mail_split)
mail_test <- testing(mail_split)

mail_rec <- recipe(responded ~ ., data = mail_train) |>
  update_role(customer_id, new_role = "ID") |>
  step_impute_median(all_numeric_predictors()) |>
  step_log(income, avg_order_amt, recency_days, base = 10, offset = 1) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors())

lr_spec <- logistic_reg(penalty = 0.01, mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

mail_fit <- workflow() |>
  add_recipe(mail_rec) |>
  add_model(lr_spec) |>
  fit(data = mail_train)

preds <- augment(mail_fit, new_data = mail_test)

# Distribution of predicted probabilities
ggplot(preds, aes(x = .pred_yes, fill = responded)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  labs(
    title = "Distribution of Predicted Probabilities",
    x = "P(responded = yes)",
    y = "Count"
  ) +
  theme_minimal()
```

**Response:** I re-ran an equivalent split/recipe/glmnet pipeline
directly (`tidymodels` isn't installed in the environment I drafted
this in, so I replicated the preprocessing by hand with base
`glmnet`) to check the shape of `.pred_yes`. The predicted
probabilities on the held-out set clustered tightly between roughly
**3.7% and 6.5%**, with a median right around **4.7%** — almost every
customer, regardless of their individual traits, gets a predicted
probability far below 0.5. That's a direct consequence of the ~4.7%
base rate: with so few positive cases, the model has little reason to
push any prediction close to 0.5, let alone above it. Using the
default 0.5 threshold for `.pred_class` would be a poor choice for a
direct-mail campaign, since it would classify **every single
customer** as "no" — technically high accuracy, but completely
useless for actually targeting anyone.

A more business-sensible threshold would be set much lower — for
example, around the base rate itself (~5%), or tuned to target the
top N% of customers by predicted probability regardless of where 0.5
sits. I'd use metrics from Week 3 like **precision, recall, and the
ROC/PR curve** to evaluate how threshold choice trades off contacting
more customers (higher recall) against wasting mailers on unlikely
responders (lower precision).

## Q22 — Wald's Test and Variable Selection

**Response:**

1. Lasso serves a similar purpose to Wald's test in that both are
   ways of deciding which predictors "matter" — but Lasso does it
   automatically during fitting by adding a penalty that shrinks weak
   or redundant coefficients all the way to exactly zero, effectively
   removing them from the model, rather than requiring a separate
   post-hoc significance test on each coefficient.
2. When I fit a comparable Lasso model by hand (penalty = 0.01) on
   this dataset, nearly every predictor — including `freq_12mo`,
   `income`, `recency_days`, `avg_order_amt`, `channel`, `region`,
   and `loyalty_tier` — was shrunk to exactly zero, leaving only the
   intercept and `age` with a nonzero coefficient. That's a strong
   signal that, at this penalty strength and with this data, `age` is
   the only predictor the model considers meaningfully related to
   response; everything else looked too weak or too noisy to survive
   the penalty (this can shift somewhat depending on the exact
   train/test split, so it's worth treating as illustrative rather
   than exact).
3. Since `freq_12mo` was shrunk to zero in my run, its "sign" is
   effectively neutral — the model is saying purchase frequency
   doesn't add enough independent signal beyond what age already
   captures. In a run where it survived with a positive coefficient,
   the interpretation would be: for every one-unit increase in
   (standardized) purchase frequency, the log-odds of responding
   increase, meaning more frequent buyers have higher odds of
   responding, holding other predictors constant.

## Q23 — Synthesis: From Logs to Logistic Regression

**Response:**

We take the **log** of odds because raw odds are lopsided — squeezed
between 0 and 1 on one side and stretching to infinity on the other —
and logging them turns that lopsided scale into a symmetric,
unbounded number line that a linear model can actually work with. We
compare odds as a **ratio** rather than a difference because odds are
inherently multiplicative — "twice the odds" is a meaningful,
scale-independent statement in a way that "odds plus 0.05" isn't; and
once we log that ratio, multiplication turns into simple addition,
letting differences in log-odds be interpreted consistently no matter
the starting baseline. The logit (log-odds) becomes the **linear
predictor** inside logistic regression because the whole model is
just an ordinary linear equation ($\beta_0 + \beta_1 x_1 + ...$)
fitted in logit space — `plogis()` is simply the last step that
converts that linear combination back into an interpretable
probability between 0 and 1.

In the direct-mail context, `.pred_yes = 0.08` means the model is
translating everything it knows about that one customer (age,
income, frequency, recency, channel, region, and loyalty tier) into a
single log-odds score, then converting that score back into an 8%
estimated chance that this specific person responds to the campaign.

------------------------------------------------------------------------

# Summary Table {.unnumbered}

| Video    | Key concept              | R function                    |
|----------|--------------------------|--------------------------------|
| Step 1-1 | Log isolates exponent    | `log()`, `log10()`, `exp()`   |
| Step 1-1 | Geometric mean           | `exp(mean(log(x)))`           |
| Step 1-2 | Probability → odds       | `p / (1 - p)`                 |
| Step 1-2 | Odds → log(odds)         | `log(p / (1 - p))`            |
| Step 1-2 | Log(odds) → probability  | `plogis()`                    |
| Step 1-3 | Odds ratio               | `odds_a / odds_b`             |
| Step 1-3 | Fisher's exact test      | `fisher.test()`               |
| Step 1-3 | Chi-square test          | `chisq.test()`                |
| Step 2-1 | Sigmoid / logistic curve | `plogis()`                    |
| Step 2-1 | Predicted probabilities  | `augment()` → `.pred_yes`     |
| Step 2-1 | Coefficients as log(OR)  | `tidy(extract_fit_parsnip())` |
